The State of Urbit:\\Eight Years After the Whitepaper

### 2024-04-10 mss.tex
Final version for proof.

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} I:1 (2024):  1–45. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Eight Years After the Whitepaper}
\fancyfoot[LE,RO]{\thepage}

@@

While the whitepaper by its own admission presented itself as a ``work-in-progress presentation'' of a ``semi-closed alpha test'', the basic claims of the Urbit whitepaper are still true today.  Most of the intervening work over the past eight years has gone into improving robustness and better support for applications.

@@

Since 2016, Nock underwent a revision from Kelvin version 5 down to 4.  This was a refinement rather than a deviation, mainly adding an opcode to manage tree mutation.  Hoon has had a number of changes (\emph{vide infra}) but is certainly recognizable as the same language.

@@

The system has primarily changed via addition, mostly growing upward in the stack: runtime and kernel infrastructure to support distributing, installing, running, and upgrading userspace applications.  Changes have been made to peer-to-peer networking, {\sc http} serving, source code management in Clay, and Arvo kernel upgrades.  Permanent network continuity was achieved at the end of 2020, largely by deleting the Ford build system vane, integrating the build system into the source control system in the Clay vane, and modifying the Arvo kernel's upgrade semantics.

@@

The other big change is the public key infrastructure ({\sc pki}).  The whitepaper's {\sc pki}, which was live from 2013 to 2019 (TODO: check date), had no double-spend protection and was effectively centralized.  To fix this, the {\sc pki} was moved to a set of Ethereum contracts.  Going on-chain made the ownership ledger for Urbit IDs censorship-resistant ... until a year later when high transaction fees on Ethereum brought network growth to a standstill and the core developers were forced to write a custom ``naive rollup'' Ethereum contract.  This brought the price of spawning a new Urbit node down from hundreds of dollars to under a dollar, at the cost of on-chain interoperability.

Going on-chain also enabled ``personal continuity breaches'', in which a ship owner can notify the network through the chain that their Urbit node was rebooted from scratch, due to data loss, unrecoverable error, or compromised security.  This feature necessitated some rework of the kernel to ensure data consistency across vanes when receiving a notification that a peer had breached.

@@

\subsection{Solid Stateness}

@@

Urbit is a solid-state interpreter, but some of its higher layers are more solid-state than others.  The scry namespace, hardly mentioned in the whitepaper, provides the primary means of preserving solid-state-ness higher up the stack, especially into userspace.

@@

A relatively early change that increased the solid-state-ness of the system was called the ``Theory of a Desk''.  This was implemented as part of the Ford Fusion project in early 2020.  Ford, the build system, was moved from being a standalone vane (kernel module) into the Clay source management vane.  When Ford stood alone, despite – or maybe exacerbated by – multiple rewrites, it was difficult to achieve full correctness of source builds during a kernel upgrade.  It was common for post-upgrade files to be built erroneously using the old kernel.  This behavior was effectively nondeterministic because it relied on the ordering of short hashes (``mug''s) that identified subscribers to Clay file-changed notifications.  Sometimes it would work correctly by luck; other times it would fail in various hard-to-reason-about ways.

@@

The next steps in increasing the utility of the scry namespace include \lstinline[style=inlinecode]{%pine} (a non-referentially-transparent ``request-at-latest'' network protocol) and ``sticky scry'', which would facilitate scry-based network subscriptions by allowing requests for unbound data to persist on a publisher until the data is published, at which point it will be pushed down to subscribers with low latency.

@@

Persistent Nock caching allows for removal of most caching-related state from the kernel and applications, allowing code to focus on business logic.  Making more logic stateless contributes strongly to solid-state-ness.  Persistent Nock caching also opens up new architectural possibilities, such as userspace build systems that share a runtime cache to avoid rerunning compilations or duplicating libraries, even if they have varying semantics.

@@

The final problem with depth-first ordering is that empirically, it is difficult to reason about.  Moves are not processed chronologically – later moves can jump to the front of the queue.  This has led to a number of subtle bugs due to later moves invalidating earlier moves, such as canceling a subscription that hadn't been established yet because the unsubscribe move jumped in front of the subscribe move.

The alternative that has been built uses breadth-first move ordering.  This is strictly chronological, does not reorder moves, has a concept of simultaneity for maintaining consistency when delivering notifications, and works equally well locally and over the network.  The major disadvantage of breadth-first ordering is that there is no automatically growing transaction: everything requires explicit acknowledgment.

The latest ``shrubbery'' experiment uses a hybrid move ordering: depth-first for local pokes within the same application, and breadth-first otherwise.

@@

``Shrubbery'' refers to an architectural concept in which all userspace data is addressible in the scry namespace.  This would include source code and application state in addition to inert file-like data.

@@

This hybrid approach can be thought of as the lowest layers of the move tree being depth-first, with breadth-first at higher layers.  Core devs expect the shrubbery experiment to be runnable within a month of this writing.

@@

Until the end of 2020, core devs would routinely conduct network-wide breaches.  Every ship on the network would need to be restarted from scratch in order to communicate with other ships.

For a user, the experience of discontinuity is one of loss of data and possibly even of friends.  This document will use the term continuity in a broad sense.

@@

The capability to perform an ``individual continuity breach'' was added.  Before going on-chain, if an Urbit suffered any major data loss, it would be permanently dead and could never communicate with peers again.  An individual continuity breach allows a user to post a transaction to chain that increments the ``rift'' number for their address, signaling to the network that the ship has been restarted from scratch.  Peer ships that were storing message sequence numbers for the breached ship reset their networking state for the breached ship.

This allows a user to continue to use the address they purchased even if their Urbit loses data.  The newly written \lstinline[style=inlinecode]{%phoenix} app implements a system of encrypted backups, so that after a breach, a user can recover application data from the latest backup.

@@

The Urbit galaxies have always been considered to form a senate, acting as the Shelling point for network governance.  Before going on chain, this was an idea.  Now it is a formalized reality.  Galaxies can vote on-chain, using a smart contract that has control over the other smart contracts.

There are two ways galaxies can vote: on (the hash of) a document to signal support of that document, and on a new constitution contract, which will trigger the old constitution contract to cede control to the new constitution contract that replaces it.  This means the galaxies can vote to upgrade the constitution as they see fit – an elegant example of social metacircularity.

@@

This was crucial for performing the last network breach at the end of 2020, establishing permanent continuity – since then, many dozens of over-the-air upgrades have been deployed.  Without a reliable upgrade process, this continuity era would be infeasible.

@@

\subsection{Continuity: Future Work}

@@

The Directed Messaging project will increase network throughput by roughly a factor of a thousand, fast enough to saturate a Gigabit ethernet link.  The Ares project will increase the maximum persistent data storage by a factor of several thousand, up to 64 Terabytes.  Ares should also improve Nock execution speed dramatically, paving the way toward Nock being able to be compiled to machine code comparable to that emitted by a C compiler.

@@

All through 2017 and 2018, Urbit would crash for the typical user on a roughly weekly basis.  Now in 2024, that frequency seems to be anecdotally more like once every few months, usually from running out of disk space due to excessive event log buildup, which has already been addressed on NativePlanet's systems, which perform automatic event log rollover and truncation.

@@

In conclusion, Urbit has grown to realize many of the promises of the 2016 whitepaper.  The system is more reliable, more scalable, more extensible, and more secure than it was then.  The network has grown to include thousands of users, and the system is now capable of supporting a much larger userbase.  While more work remains to be done, the kernel of Urbit is sound and the contours of the kelvin-cooled system have begun to assume more resolution.  \tombstone{}

\section{Appendix}

This appendix addresses developments from the whitepaper by section.  While this is somewhat narratively unsatisfactory, it provides an easy reference.  In cases where no significant changes have been made, the section has been elided.

@@

Web2 interoperability (TODO: quote the whitepaper) has been partially superseded by Web3 interoperability.  In 2016 Web3 barely existed, but in 2024 it's a relatively established industry.  Over the last two years there's been broadening acceptance of Urbit's complementary nature to blockchains.

@@

\begin{enumerate}
  \item  The Nock 5 equality check opcode was tweaked to enforce taking two subexpressions, rather than optionally taking a single expression that could return a cell (pair), whose head and tail would then be compared for equality.  This optionality turned out to be annoying when writing an interpreter, and it provided little utility.
  \item  Nock version 4 added a new Nock 10 opcode to replace part of a tree with a new value.  Creating mutant trees like this is a basic operation that Hoon code performs regularly, using the \lstinline[style=inlinecode]{%=} rune (TODO: should that be centar?) and its variants (\lstinline[style=inlinecode]{%*}, \lstinline[style=inlinecode]{%_}, \lstinline[style=inlinecode]{=.}, and \lstinline[style=inlinecode]{=:}).  Before the addition of the Nock 10 opcode, the Hoon compiler would use the type of the trees involved in a mutation expression to codegen a tuple of Nock ``auto-cons'' expressions (cells of expressions that return cells of their results) that would reconstruct the mutant tree out of the new subtree cons'd together with the remaining pieces of the original tree.

@@

The Udon and Sail domain-specific languages were added to Hoon, for markdown-like rich text and XML/HTML, respectively.  There is a tentative consensus among core developers that these DSLs should be kicked up a layer out of the base Hoon language, but making Hoon syntax recursively extensible likely requires splitting out the definition of the Hoon AST into a recursion scheme, which is nontrivial.

@@

Wetness in general is an interesting and novel approach to both polymorphic and recursive types, both of which are directly represented as lazily evaluated types (Urbit core devs often call the ``evaluation'' step ``type inference'' informally, but it corresponds to the ``type synthesis'' direction of a bidirectional typechecker in type theory terminology).  This has pros and cons.

@@

There is some interest among core devs in experimenting with a more constrained ``type scheme'' (TODO: check with ~fodwyt for correct term) that could be compiled to a fully evaluated data structure with De Brouijn indices (or maybe De Brouijn levels), in the hopes that could make Hoon's type system easier to learn and reason about.

The formal Arvo boot sequence as described in the whitepaper was not implemented at the time it was written.  It was deployed in 2019.  Around the same time a lot of other runtime changes were made.  The runtime was split into an I/O process and an Arvo worker process.  The event log was switched from a custom binary file to the {\sc lmdb} database.  TODO what else happened around that time?

@@

For example, Ames uses a synthetic duct for each ``flow'' (Ames's version of a socket connection) to set packet re-send timers for that flow and pass requests to other vanes.  This disconnects the flow's duct from the duct that the runtime used to inject the event into Arvo, preventing Unix process restarts from causing desynchronization.

As another example, Gall now uses a per-agent synthetic duct for the moves coming out of that agent.  Note that this approach has a known flaw, where if an agent sets a timer in response to hearing a request from another ship, it needs to remember that ship in order to obtain the same duct to cancel the timer.  XX this doesn't make sense – how does the timer get into the wire?

The \lstinline[style=inlinecode]{.^} dotket rune no longer blocks until complete.  Because the agent model expects non-blocking I/O, a single dotket could make an entire agent stuck if it can't be completed.  While a lot of work has gone into fleshing out the scry namespace, core devs are now entertaining the notion that the expectation of synchronous access to the namespace might not be worth it for agents.  An Urbit Improvement Proposal has been submitted that would remove the dotket rune in favor of a move-based interface to the namespace.

@@

The whitepaper says ``most data is now in Clay''.  This is no longer true.  Now most data lives in Gall, in agents.  Clay has proven difficult to use for storing application data for a number of reasons, and Gall has become more featureful.  There is consensus among the core devs that a better userspace model should be devised to unify Clay and Gall.  One such model, called ``shrubbery'', is now being investigated.

@@

Ames is still sequential and idempotent for commands.  Despite having been described as ``content-centric'' and ``source-independent'' in the whitepaper, those two properties are only now being realized, in the ``directed messaging'' project that is currently underway.  Directed messaging refactors Ames to be much more like the Named Data Networking project: packets and messages are both constrained to be request/response, with routing being achieved using a Pending Interest Table just like in {\sc ndn}.

@@

The Ares project has another new approach to jet matching, which is to resolve the matching statically during a transpilation step from Nock to a more optimized representation.  This is facilitated by a novel static analysis technique for Nock, called Subject Knowledge Analysis ({\sc ska}), which analyzes the data dependencies within a Nock formula, allowing inference about the tree shape of the ``subject'' (environment, or scope) at different points of the computation.  {\sc ska} should allow for more efficient registerization of intermediate values, and will also provide enough information to statically compute whether a jet will match.  The goal is for Ares to convert a hot loop in Nock into similar machine code that would be emitted by a C compiler, including direct jumps back to the start of the loop.

@@

An interesting optimization that was added to Vere more recently is called ``tail-call modulo cons'', allowing recursive loops that are almost tail-recursive to be optimized as if they are tail-recursive, as long as the only operation that happens after each recursive call is a ``cons'' that prepends or appends another value to the result of the recursion.
TODO link to paper about tail-call modulo cons

@@

Hoon has grown, but the semantics have largely remained intact.  \lstinline[style=inlinecode]{$span} and \lstinline[style=inlinecode]{$twig} have been renamed, but this may be reverted at some point.  (Most core devs prefer the older names.)  Core types have ``chapters'' added to them, mostly for documentation purposes.

@@

\begin{enumerate}
  \item  \lstinline[style=inlinecode]{!<} zapgal
  \item  \lstinline[style=inlinecode]{;<} micgal
  \item  \lstinline[style=inlinecode]{|@} barpat
  \item  \lstinline[style=inlinecode]{|$} barbuc
  \item  \lstinline[style=inlinecode]{#/} haxfas TODO
  \item  \lstinline[style=inlinecode]{?#} wuthax (still experimental)
  \item  \lstinline[style=inlinecode]{+*} lustar
  \item  \lstinline[style=inlinecode]{+$} lusbuc
  \item  \lstinline[style=inlinecode]{$~} bucsig
  \item  \lstinline[style=inlinecode]{^~} tissig
  \item  \lstinline[style=inlinecode]{$|} bucbar
  \item  \lstinline[style=inlinecode]{$>} bucgar
  \item  \lstinline[style=inlinecode]{$<} bucgal
  \item  \lstinline[style=inlinecode]{$:} bucmic
\end{enumerate}

@@

The Hoon type system used to use an idempotence check when coercing a raw noun into a typed noun.  XX why?  As of the 2019 type system revision, that check has been removed, and the mold gate will crash (\lstinline[style=inlinecode]{!!} zapzap) if the noun fails to validate.

@@

This represents a shift in thinking among core devs.  Rather than a more monolithic, static, trusted userspace in earlier designs, userspace is now thought of as more multicentric, dynamic, and untrusted.  If apps can all be trusted, there is no need to protect one app against the others.  In practice, however, there is a compelling need to prevent a single malicious (or, equivalently, badly written) app from breaking the security properties of any other app.  Allowing an app to see which other local app is poking it provides enough information to resolve a major class of confused deputy problems without implementing a full-fledged capabilities system.

@@

With the Directed Messaging rewrite, a new packet authentication protocol has been added, called LockStep.  This protocol is based on the {\sc bao} streaming authentication protocol.  Like {\sc bao}, it uses the Blake3 tree hash to establish fingerprints for both individual packets and the full message datum.  It differs from {\sc bao} in two main ways: it is packet-oriented rather than byte-oriented, and it guarantees that as long as packets are received in order, each packet can be authenticated immediately without waiting for any other packets.

@@

\paragraph{9.9 Packet Rejection}  Cryptography failures (failure to decrypt or authenticate a packet) still cause a silent packet drop.  TODO what was the term that wireguard uses for this security property?  To streamline the implementation of this, the \lstinline[style=inlinecode]{bail:evil} error tag was added to the system so the runtime can know not to inject an error notification event into Arvo in the case of cryptographic failure.

@@

Deferred acknowledgments are still vaporware.

@@

The Urbit roadmap is available at \url{https://roadmap.urbit.org/}.

\#\# Section 12 Related Work

XX have we learned from any other projects since the whitepaper?  {\sc ndn}, perhaps...

\#\# Remaining

XX turn into prose

- scry over {\sc http}
- eauth
- {\sc tls} termination, letsencrypt
- pill management
- aqua, pH, +test, -test
- devstream process
- published roadmap
- {\sc uip}s
- UF leadership
- \%landscape, \%glob
- Clay global cache
- reliability and usability of Tlon chat
- reliability and performance of webpage loads
- network scale improvements
- web terminal
- pointer compression
- \%bout timing hints
- color printing
- ordered map data structure
- compile-time evaluation in Hoon
- zero-click maintenance work
- Bridge
- |fuse
- scry arc, subscription reform
- shrubbery
- urWASM, Luau, Acacia
- abortive in-Hoon compiler caching
- Lagoon and MNIST

- sections
  - Early vane explosion, 2015-2017
  - Hoon, Ford, Eyre, \& Vere rewrites, 2017-2019
  - Azimuth \& Ames rewrite, 2018-2019
  - bytecode interpreter, 2028-2019
  - Ford Fusion \& continuity, 2020
  - security audit on Ames, 2020
  - layer2, 2021
  - software distribution, 2021
    - maturation, 2022-2024
      - Clay tombstoning, 2021-2022
      - Clay streamlining, 2022
      - Clay referentially transparent caching, 2022
      - "agent-in-clay" upgrade overhaul, 2022
      - essential desks, 2024
  - network scaling, 2023-2024
    - remote scry protocols
    - pinging improvements
    - dead flow consolidation
    - directed messaging
  - runtime improvements, ongoing
    - |mass, |pack, |meld, |trim
    - .run, next
    - .run chop, roll (epochs)
  - Ares, 2022-2024
  - upcoming
    - shrubbery
    - media streaming
    - numerics
    - other languages: WASM, Lua, Acacia
    - security
    - hoon type system improvements
    - unikernel
    - NockPU

>>>>

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} I:1 (2024):  1–46. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Eight Years After the Whitepaper}
\fancyfoot[LE,RO]{\thepage}

@@

While the whitepaper by its own admission presented itself as a ``work-in-progress presentation'' of a ``semi-closed alpha test'', the basic claims of the Urbit whitepaper are still true today.  Most of the intervening work over the past eight years has gone into improving robustness and better support for applications.  A capsule history of ``late early'' Urbit includes the following milestones:
\begin{itemize}
  \item  2015–2017:  Early vane explosion
  \item  2017–2019:  Hoon, Ford, Eyre, and Vere rewrites refine the system
  \item  2018–2019:  Azimuth and Ames rewrite refine the network
  \item  2018–2019:  Nock bytecode interpreter
  \item  2020:  Ford Fusion enables network continuity
  \item  2020:  Security audit on Ames
  \item  2021:  Azimuth Layer2
  \item  2021:  Software distribution introduced
  \item  2022–2024:  Software distribution maturation
  \item  2023–2024:  Network scaling
\end{itemize}

\noindent
This article will discuss the changes to Urbit since the whitepaper, focusing on the runtime, the network, the kernel, and the userspace.  It will also discuss the current state of the project and intended refinements.

@@

Since 2016, Nock underwent a revision from Kelvin version 5K down to 4K.  This was a refinement rather than a deviation, mainly adding an opcode to manage tree mutation.  Hoon has undergone a number of changes but is certainly recognizable as the same language.

@@

\sloppy
The system has primarily changed via addition, mostly growing upward in the stack: runtime and kernel infrastructure to support distributing, installing, running, and upgrading user\-space applications.  Changes have been made to peer-to-peer networking, {\sc http} serving, source code management in Clay, and Arvo kernel upgrades.  Permanent network continuity was achieved at the end of 2020, largely by deleting the Ford build system vane, integrating the build system into the source control system in the Clay vane, and modifying the Arvo kernel's upgrade semantics.

@@

The other big change is the public key infrastructure ({\sc pki}).  The whitepaper's {\sc pki}, which was effectively live from 2013 to 2019, had no double-spend protection and was effectively centralized.  To fix this, the {\sc pki} was moved to a set of Ethereum contracts.  Going on-chain made the ownership ledger for Urbit IDs censorship-resistant ... until a year later when high transaction fees on Ethereum brought network growth to a standstill and the core developers were forced to write a custom ``naive rollup'' Ethereum contract.  This brought the price of spawning a new Urbit node down from hundreds of dollars to under a dollar, at the cost of on-chain interoperability.

\sloppy
Going on-chain also enabled ``personal continuity breaches'', in which a ship owner can notify the network through the chain that their Urbit node was rebooted from scratch, due to data loss, unrecoverable error, or compromised security.  This feature necessitated some rework of the kernel to ensure data consistency across vanes when receiving a notification that a peer had breached.

@@

\subsection{Solid-Stateness}

@@

Urbit is a solid-state interpreter but some of its higher layers are more solid-state than others.  The scry namespace, hard\-ly mentioned in the whitepaper, provides the primary means of preserving solid-stateness higher up the stack, especially into userspace.

@@

A relatively early change that increased the solid-stateness of the system was called the ``Theory of a Desk''.  This was implemented as part of the Ford Fusion project in early 2020.  Ford, the build system, was moved from being a standalone vane (kernel module) into the Clay source management vane.  When Ford stood alone, despite – or maybe exacerbated by – multiple rewrites, it was difficult to achieve full correctness of source builds during a kernel upgrade.  It was common for post-upgrade files to be built erroneously using the old kernel.  This behavior was effectively nondeterministic because it relied on the ordering of short hashes (``mug''s) that identified subscribers to Clay file-changed notifications.  Sometimes it would work correctly by luck; other times it would fail in various hard-to-reason-about ways.

@@

\sloppy
The next steps in increasing the utility of the scry namespace include \lstinline[style=inlinecode]{%pine} (a non-referentially-transparent ``request-at-latest'' network protocol) and ``sticky scry'', which would facilitate scry-based network subscriptions by allowing requests for unbound data to persist on a publisher until the data is published, at which point it will be pushed down to subscribers with low latency.

@@

Persistent Nock caching allows for removal of most caching-related state from the kernel and applications, allowing code to focus on business logic.  Making more logic stateless contributes strongly to solid-stateness.  Persistent Nock caching also opens up new architectural possibilities, such as userspace build systems that share a runtime cache to avoid rerunning compilations or duplicating libraries, even if they have varying semantics.

@@

The final problem with depth-first ordering is that it is empirically difficult to reason about.  Moves are not processed chronologically – later moves can jump to the front of the queue.  This has led to a number of subtle bugs due to later moves invalidating earlier moves, such as canceling a subscription that hadn't been established yet because the unsubscribe move jumped in front of the subscribe move.

An alternative has been built which uses breadth-first move ordering.  This is strictly chronological, does not reorder moves, has a concept of simultaneity for maintaining consistency when delivering notifications, and works equally well locally and over the network.  The major disadvantage of breadth-first ordering is that there is no automatically growing transaction: everything requires explicit acknowledgment.

The latest ``shrubbery'' (q.v.) experiment uses a hybrid move ordering: depth-first for local pokes within the same application, and breadth-first otherwise.

@@

``Shrubbery'' refers to an architectural concept in which all user\-space data is addressible in the scry namespace.  This would include source code and application state in addition to inert file-like data.

@@

This hybrid approach can be thought of as the lowest layers of the move tree being depth-first, with breadth-first at higher layers.  Core developers expect the shrubbery experiment to be runnable within a month of this writing.

@@

Until the end of 2020, core developers would routinely conduct network-wide breaches.  Every ship on the network would need to be restarted from scratch in order to communicate with other ships.

For a user, the experience of discontinuity is one of loss of data and possibly even of friends.  This document will use the term ``continuity'' in a broad sense.

@@

\sloppy
The capability to perform an ``individual continuity breach'' was added.  Before going on-chain, if an Urbit suffered any major data loss, it would be permanently dead and could never communicate with peers again.  An individual continuity breach allows a user to post a transaction to chain that increments the ``rift'' number for their address, signaling to the network that the ship has been restarted from scratch.  Peer ships that were storing message sequence numbers for the breached ship reset their networking state for the breached ship.

\sloppy
This allows a user to continue to use the address they purchased even if their Urbit loses data.  The newly written \lstinline[style=inlinecode]{%phoenix} app implements a system of encrypted backups, so that after a breach, a user can recover application data from the latest backup.

@@

\sloppy
The Urbit galaxies have always been considered to form a senate, acting as the Shelling point for network governance.  Before going on chain, this was merely an idea – now it is a formalized reality.  Galaxies can vote on-chain, using a smart contract that has control over the other smart contracts.

There are two ways galaxies can vote: on (the hash of) a document to signal support of that document, and on a new constitution contract, which will trigger the old constitution contract to cede control to the new constitution contract that replaces it.  This means the galaxies can vote to upgrade the constitution as they see fit – an elegant example of social meta\-circularity.

@@

\sloppy
This was crucial for performing the last network breach at the end of 2020, establishing permanent continuity – since then, many dozens of over-the-air upgrades have been deployed.  Without a reliable upgrade process, this continuity era would be infeasible.

@@

\subsection{Future Work}

@@

The Directed Messaging project will increase network throughput by roughly a factor of a thousand, fast enough to saturate a gigabit ethernet link.  The Ares project will increase the maximum persistent data storage by a factor of several thousand, up to 64 Terabytes.  Ares should also improve Nock execution speed dramatically, paving the way toward Nock being able to be compiled to machine code comparable to that emitted by a C compiler.

@@

All through 2017 and 2018, Urbit would crash for the typical user on a roughly weekly basis.  As of 2024, that frequency seems to be anecdotally once every few months, usually from running out of disk space due to excessive event log buildup; this has already been addressed on NativePlanet's systems, which perform automatic event log rollover and truncation.

@@

Urbit development has begun to realize many of the promises of the 2016 whitepaper.  The system is more reliable, more scalable, more extensible, and more secure than it was then.  The network has grown to include thousands of users, and the system is now capable of supporting a much larger userbase.  Indeed, this article omits many quality-of-life improvements, such as scrying over {\sc http}, EAuth, \lstinline[style=inlinecode]{%bout} timing hints, bootstrap pill management, reliability and usability of Tlon's social media application, color printing, ordered map data structures, compile-time evaluation in Hoon, the Bridge contract operations tool for Azimuth and details of the Urbit HD wallet, the 
The published roadmap (available at \url{https://roadmap.urbit.org/}) provides more detail on past and planned accomplishments at a resolution impossible in a single article.

While more work remains to be done, the kernel of Urbit is sound and the contours of the kelvin-cooled system have begun to assume more resolution.  \tombstone{}

\section*{Appendix}
\addcontentsline{toc}{section}{Appendix}

This appendix addresses developments from content in the whitepaper by section.  While this is somewhat narratively unsatisfactory, it provides an easy reference.  In cases where no significant changes have been made, the section has been omitted.

@@

Web 2.0 interoperability (as the whitepaper discusses, Urbit as a ``browser for the server side'') has been partially superseded by Web3 interoperability.  In 2016 Web3 barely existed, but by 2024 it has become a relatively established industry.  Over the last few years, acceptance of Urbit's complementary nature to blockchains has broadened.

@@

\begin{enumerate}
  \item  The Nock 5 equality check opcode was tweaked to enforce taking two subexpressions, rather than optionally taking a single expression that could return a cell (pair), whose head and tail would then be compared for equality.  This optionality turned out to be annoying when writing an interpreter, and it provided little utility.
  \item  Nock version 4 added a new Nock 10 opcode to replace part of a tree with a new value.  Creating mutant trees like this is a basic operation that Hoon code performs regularly, using the \lstinline[style=inlinecode]{%=} rune and its variants (\lstinline[style=inlinecode]{%*}, \lstinline[style=inlinecode]{%_}, \lstinline[style=inlinecode]{=.}, and \lstinline[style=inlinecode]{=:}).  Before the addition of the Nock 10 opcode, the Hoon compiler would use the type of the trees involved in a mutation expression to codegen a tuple of Nock ``auto-cons'' expressions (cells of expressions that return cells of their results) that would reconstruct the mutant tree out of the new subtree cons'd together with the remaining pieces of the original tree.

@@

The Udon and Sail domain-specific languages were added to Hoon, for Markdown-like rich text and {\sc xml}/{\sc html}, respectively.  There is a tentative consensus among core developers that these {\sc dsl}s should be kicked up a layer out of the base Hoon language, but making Hoon syntax recursively extensible likely requires splitting out the definition of the Hoon {\sc ast} into a recursion scheme, which is nontrivial.

@@

Wetness in general is an interesting and novel approach to both polymorphic and recursive types, both of which are directly represented as lazily evaluated types (Urbit core developers often call the ``evaluation'' step ``type inference'' informally, but it corresponds to the ``type synthesis'' direction of a bidirectional typechecker in type theory terminology).  This has pros and cons.

@@

There is some interest among core developers in experimenting with a more constrained type scheme that could be compiled to a fully evaluated data structure with de Bruijn indices.  Such a quantified type scheme with binders could make Hoon's type system easier to learn and reason about.

The formal Arvo boot sequence as described in the whitepaper was not implemented at the time it was written.  It was deployed in 2019.  Around the same time a lot of other runtime changes were made.  The runtime was split into an I/O process and an Arvo worker process.  The event log was switched from a custom binary file to the {\sc lmdb} database.

@@

For example, Ames uses a synthetic duct for each ``flow'' (Ames's version of a socket connection) to set packet re-send timers for that flow and pass requests to other vanes.  This disconnects the flow's duct from the duct that the runtime used to inject the event into Arvo, preventing Unix process restarts from causing desynchronization.  As another example, Gall now uses a per-agent synthetic duct for the moves coming out of that agent.\footnote{Note that this approach currently has a known flaw, wherein if an agent sets a timer in response to hearing a request from another ship, it needs to remember that ship in order to obtain the same duct to cancel the timer, cf. \href{https://github.com/urbit/urbit/issues/6884}{issue \#6884}.}

The \lstinline[style=inlinecode]{.^} dotket rune no longer blocks until complete.  Because the agent model expects non-blocking I/O, a single dotket could make an entire agent stuck if it can't be completed.  While a lot of work has gone into fleshing out the scry namespace, core developers are now entertaining the notion that the expectation of synchronous access to the namespace might not be worth it for agents.  An Urbit Improvement Proposal has been submitted that would remove the \lstinline[style=inlinecode]{.^} dotket rune in favor of a move-based interface to the namespace (see pp. TODO in this issue).

@@

The whitepaper states that ``most data is now in Clay''.  This is no longer true.  Now most data lives in Gall as agents and their state.  Clay has proven difficult to use for storing application data for a number of reasons, and Gall has become more featureful.  There is consensus among the core developers that a better userspace model should be devised to unify Clay and Gall.  One such model, called ``shrubbery'', is now being investigated.

@@

Ames is still sequential and idempotent for commands.  Despite having been described as ``content-centric'' and ``source-independent'' in the whitepaper, those two properties are only now being realized, in the ``directed messaging'' project that is currently underway.  Directed messaging refactors Ames to be much more like the Named Data Networking project \citep{Zhang2014}: packets and messages are both constrained to be request/response, with routing being achieved using a Pending Interest Table just like in {\sc ndn}.

@@

The Ares project has another new approach to jet matching, which is to resolve the matching statically during a transpilation step from Nock to a more optimized representation.  This is facilitated by a novel static analysis technique for Nock, called Subject Knowledge Analysis ({\sc ska}), the brainchild of \patp{ritpub-sipsyl}.  {\sc ska} analyzes the data dependencies within a Nock formula, allowing inference about the tree shape of the ``subject'' (environment, or scope) at different points of the computation.  {\sc ska} should allow for more efficient registerization of intermediate values, and will also provide enough information to statically compute whether a jet will match.  The goal is for Ares to convert a hot loop in Nock into similar machine code that would be emitted by a C compiler, including direct jumps back to the start of the loop.

@@

An interesting optimization that was added to Vere more recently is called ``tail-call modulo cons'' \citep{Leijen2023}.  This approach allows recursive loops that are almost tail-recursive to be optimized as if they are tail-recursive, as long as the only operation that happens after each recursive call is a ``cons'' that prepends or appends another value to the result of the recursion.

@@

Hoon has grown, but the semantics have largely remained intact.  \lstinline[style=inlinecode]{$span} and \lstinline[style=inlinecode]{$twig} have been renamed, but this may be reverted at some point.  (Most core developers prefer the older names.)  Core types have ``chapters'' added to them, mostly for documentation purposes.

@@

\begin{enumerate}
  \item  \lstinline[style=inlinecode]{|$} barbuc (produce mold builder wet gate)
  \item  \lstinline[style=inlinecode]{|@} barpat (produce wet core)
  \item  \lstinline[style=inlinecode]{$|} bucbar (produce structure to satisfy validator)
  \item  \lstinline[style=inlinecode]{$<} bucgal (filter a mold, exclude)
  \item  \lstinline[style=inlinecode]{$>} bucgar (filter a mold, match)
  \item  \lstinline[style=inlinecode]{$:} bucmic (enter manual value mode)
  \item  \lstinline[style=inlinecode]{$~} bucsig (define default custom type value)
  \item  \lstinline[style=inlinecode]{#/} haxfas (produce typed path)
  \item  \lstinline[style=inlinecode]{+$} lusbuc (type constructor arm)
  \item  \lstinline[style=inlinecode]{+*} lustar (door alias)
  \item  \lstinline[style=inlinecode]{;<} micgal (monadic do notation)
  \item  \lstinline[style=inlinecode]{^~} tissig (compose many expressions)
  \item  \lstinline[style=inlinecode]{?#} wuthax (match type, experimental)
  \item  \lstinline[style=inlinecode]{!<} zapgal (extract vase to mold)
\end{enumerate}

@@

The Hoon type system used to use an idempotence check when coercing a raw noun into a typed noun.  As of the 2019 type system revision, that check has been removed, and the mold gate will crash (\lstinline[style=inlinecode]{!!} zapzap) if the noun fails to validate.

@@

This represents a shift in thinking among core developers.  Rather than a more monolithic, static, trusted userspace in earlier designs, userspace is now thought of as more multicentric, dynamic, and untrusted.  If apps can all be trusted, there is no need to protect one app against the others.  In practice, however, there is a compelling need to prevent a single malicious (or, equivalently, badly written) app from breaking the security properties of any other app.  Allowing an app to see which other local app is poking it provides enough information to resolve a major class of confused deputy problems without implementing a full-fledged capabilities system.

@@

With the Directed Messaging rewrite, a new packet authentication protocol has been added, called LockStep.  This protocol is based on the {\sc bao} streaming authentication protocol \citep{BAO}.  Like {\sc bao}, it uses the Blake3 tree hash to establish fingerprints for both individual packets and the full message datum.  It differs from {\sc bao} in two main ways: it is packet-oriented rather than byte-oriented, and it guarantees that as long as packets are received in order, each packet can be authenticated immediately without waiting for any other packets.

@@

\paragraph{9.9 Packet Rejection}  Cryptography failures (failure to decrypt or authenticate a packet) still cause a silent packet drop.  To streamline the implementation of this, the \lstinline[style=inlinecode]{bail:evil} error tag was added to the system so the runtime can know not to inject an error notification event into Arvo in the case of cryptographic failure.

@@

Deferred acknowledgments remain vaporware.

@@
